---
title: "Regression GLS"
author: "Luke Ghallahorne"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: yes
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning = FALSE, message = FALSE}
library(automap)
library(gstat)
library(nlme)
library(spdep)
library(ncf)
library(tidyverse)
library(spatialreg)
```

# Data
```{r}
birds <- read.csv("data/birdDiv.csv")
head(birds)

birds_sf <- st_as_sf(birds, coords = c("UTME", "UTMN"), crs = 32610)
birds_sf
```

```{r}
ggplot() +
  geom_sf(data = birds_sf, aes(col = birdDiv, size = birdDiv, alpha = 0.95)) +
  scale_color_continuous(type = "viridis") +
  guides(size = "none", alpha = "none")

ggplot() +
  geom_sf(data = birds_sf, aes(col = plantDiv, size = plantDiv, alpha = 0.95)) +
  scale_color_continuous(type = "viridis") +
  guides(size = "none", alpha = "none")
```

# Analysis
## Naive Model

```{r}
# fit model assuming no autocorrelation in residuals with gls()
glsNaive <- gls(birdDiv ~ plantDiv, birds)
summary(glsNaive)
```

The initial model of bird diversity as a function of plant diversity gives a positive slope that is significantly different from zero (slope = 1.136, t = 5.445, p < 0.001) and has an error of 0.2087. However, if there is autocorrelation in the residuals, the OLS linear model may not be efficiently predicting values (and overestimating the t-value for plantDiv).

## Visualize Residuals

```{r}
# add model residuals to dataframe
birds$glsNaiveResids <- residuals(glsNaive, type = "normalized")
birds_sf <- st_as_sf(birds, coords = c("UTME", "UTMN"), crs = 32610)
birds_sf
```

```{r}
ggplot() +
  geom_sf(data = birds_sf, aes(col = glsNaiveResids, size = glsNaiveResids,
                               alpha = 0.95)) +
  scale_color_continuous(type = "viridis") +
  guides(size = "none", alpha = "none")
```

Residuals from the model look very strongly autocorrelated, with low error values clumped to the southwest and higher error clumped in the northeast.

### Correlogram
```{r}
# test for spatial autocorrelation using a correlogram
residsI <- spline.correlog(x = birds$UTME, y = birds$UTMN,
                           z = birds$glsNaiveResids, 
                           resamp = 50, quiet = TRUE)
# save coordinate points as matrix for dist()
points <- cbind(birds$UTME, birds$UTMN)

# plot Moran's I for residuals
plot(residsI, xlim = c(0, max(dist(points))/3))
```

The correlogram of Moran's I shows that there is strong autocorrelation in the residuals at distances below ~ 5000m. To account for this, I will use a variogram model of the residuals to better inform the bird diversity model.

## Variogram Fitting

```{r}
# using autofitVariogram() to produce experimental and fitted variogram
## tests Gaussian, Spherical, and Exponential fits and produces the best
birds_fit <- autofitVariogram(glsNaiveResids~1, input_data = birds_sf, 
                      model = c("Gau", "Sph", "Exp"))
plot(birds_fit)

```

The best-fit variogram model uses a Gaussian framework (nugget = 0.05, sill = 0.83, range = 2398). This will be used as a correlation structure to find the variance-covariance matrix for the updated model.

### Cross Validation
```{r}
# observational variogram
birds_var <- variogram(glsNaiveResids~1, birds_sf)
# Variogram model
birds_naivemod <- vgm(psill = 0.83, model = "Gau", range = 2398, nugget = 0.05)
# fit variogram model
birds_fit <- fit.variogram(object = birds_var, model = birds_naivemod)

# k-fold cross validation
## I believe the krige.cv() function should be appropriate here
set.seed(42)
n <- 10
birds_kfold_sf <- krige.cv(formula = glsNaiveResids ~ 1,
                               locations = birds_sf,
                               model = birds_fit,
                               nfold = n,
                               verbose = FALSE)
birds_kfold_sf
```

```{r}
birds_kfold <- data.frame(birds_kfold_sf$observed, birds_kfold_sf$var1.pred, birds_kfold_sf$fold)
colnames(birds_kfold) <- c("observed", "var1.pred", "fold")
head(birds_kfold)

rsq <- 1:n
rmse <- 1:n
fold <- 1:n

mod_fit <- data.frame(rsq, rmse, fold)

for (i in 1:n) {
  birds_kfold_sub <- subset(birds_kfold, fold == i)
  
  mod_fit$rsq[i] <- cor(birds_kfold_sub$observed, birds_kfold_sub$var1.pred)^2
  
  mod_fit$rmse[i] <- sqrt(mean((birds_kfold_sub$observed - birds_kfold_sub$var1.pred)^2))
}
mod_fit

mean(mod_fit$rsq)
mean(mod_fit$rmse)
```

The variogram model is not a very good fit of the residuals ($R^2$ = 0.336, RMSE = 0.919).

## Updated Model
```{r}
csSpatial <- corSpatial(form=~UTME+UTMN, nugget = TRUE, type = "gaussian")
glsUpdated <- update(glsNaive, correlation = csSpatial)
summary(glsUpdated)
```

The updated regression model shows a different result to the naive model: a non-significant slope in plant diversity predicting bird diversity.

```{r}
# add the residuals to the spatial dat object
birds$glsResids <- residuals(glsUpdated,type="normalized")
birds_sf <- st_as_sf(birds, coords = c("UTME", "UTMN"), crs = 32610)

# Map the residuals from gls
ggplot() +
  geom_sf(data = birds_sf, aes(col = glsResids, size = glsResids,
                               alpha = 0.95)) +
  scale_color_continuous(type = "viridis") +
  guides(size = "none", alpha = "none")
```

The residuals of the updated model look significantly less autocorrelated than the naive model's. 

```{r}
# test for spatial autocorrelation using a correlogram
residsI <- spline.correlog(x = birds$UTME, y = birds$UTMN,
                           z = birds$glsResids, 
                           resamp = 50, quiet = TRUE)
# save coordinate points as matrix for dist()
points <- cbind(birds$UTME, birds$UTMN)

# plot Moran's I for residuals
plot(residsI, xlim = c(0, max(dist(points))/3))
```

Moran's I of the residuals shows that there is no significant spatial autocorrelation among them.

# Interpretation.
Not accounting for spatial autocorrelation, plant diversity appears to be a moderate predictor of bird diversity. However, when we adjust for autocorrelation, the slope is no longer significant. Other factors contributing to the spatial autocorrelation between plant and bird diversity - possibly climate/weather, terrain/topography, or other environmental variables not included in the data - are likely responsible for this, as bird diversity alone is not a significant predictor of plant diversity. 




