---
title: "A7 Time Series"
author: "Luke Ghallahorne"
date: "`r Sys.Date()`"
output:   
  html_document:
    toc: true
    toc_depth: 3
    toc_float: yes
    theme: flatly
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# library
library(tidyverse)
# lubridate 1.9.3 now included in tidyverse 2.0.0
library(PNWColors)
```

```{r}
# color palettes
## save color palette, depending on number of observations (n)
### if using a continuous scale, can have n much higher than # observations
pal1 <- pnw_palette("Sunset2", n = 399)
pal2 <- pnw_palette("Starfish", n = 4)
# view color palette:
#pal1
# see name options for palette generation:
names(pnw_palettes)
```


# Weather Data

These data consist of monthly weather data from the Bellingham airport weather station from January 2000 to present. 

```{r}
# load weather data (RDS file)
## using an easier variable name since I am only working with one dataset
weather <- readRDS("kbliMonthlyTempPrcp.rds")
head(weather)
str(weather)
class(weather$DATE)
```

## Pull Year and Month

```{r}
weather <- weather %>%
  mutate(YEAR = year(DATE),
         MONTH = month(DATE))
head(weather)
```

## Pull Month Name
```{r}
weather <- weather %>%
  mutate(MONTHNAME = month(DATE, label = TRUE))
head(weather)
```

## Pivoting
```{r}
weather_long <- weather %>%
  select(DATE, TEMP, PRCP) %>%
  pivot_longer(cols = c(TEMP, PRCP))
```

## Aggregating Data
```{r}
weather_AnnualTemp <- weather %>%
  filter(YEAR < 2024) %>%
  group_by(YEAR) %>%
  summarize(TEMP = mean(TEMP))
head(weather_AnnualTemp)
```

# Plot 1: Average Annual Summer Temperatures

## Data Wrangling

```{r}
AvgSummerTemp <- weather %>%
  # filter for summer months: June (6), July (7), and August (8)
  filter(MONTH %in% 6:8) %>%
  # group by year to take summer averages
  group_by(YEAR) %>%
  # summarize each year using mean() 
  summarize(TEMP = mean(TEMP)) %>%
  # choose only columns of interest for this plot 
  select(YEAR, TEMP)
```

## Plot

```{r}
plot_AvgSummerTemp <- ggplot(data = AvgSummerTemp) +
    ## data was reordered so that the color attributes would align
  # add connecting line, colored by same palette
  geom_line(aes(x = YEAR, y = TEMP, color = TEMP)) +
  # add points for each year, colored by temp
  geom_point(aes(x = YEAR, y = TEMP, color = TEMP)) +
  # change axis labels
  ## use \u00b0 in character string to get degree symbol
  # change color palette
  scale_color_gradientn(colors = pal1) +
  labs(x = "Year", y = "Temperature (\u00b0C)",
       title = "Average Summer Temperature (\u00b0C)",
       subtitle = "2000 - 2024 | Bellingham International Airport Weather Station") +
  # change legend title
  ## use \n to add linebreak to character string
  guides(color = guide_legend(title =  "Temp \n (\u00b0C)")) +
  # center plot title and subtitle
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
plot_AvgSummerTemp
```

Fig 1: Average summer temperature recorded at Bellingham International Airport's weather station from 2000 to 2024. Temperatures were averaged across June, July, and August for each year.


# Plot 2: Cumulative Precipitation by Year

## Data Wrangling

```{r}
CumPrcp <- weather %>%
  # select only Year and Precipitation columns
  select(YEAR, PRCP) %>%
  # group by year 
  group_by(YEAR) %>%
  # warnings said to use reframe instead of summarize - not entirely sure why, since CumPrcp has the same number of rows as weather
  # cumsum() creates a vector that is the cumulative sum of the values
  reframe(sumPRCP = cumsum(PRCP)) %>%
  # add month back into the dataframe
  add_column(MONTH = weather$MONTH,
             DATE = weather$DATE) %>%
  # add quarter column for seasons
  mutate(SEASON = as.factor(quarter(DATE)))

# create reference dataframe to add Season Names to DF
seasons <- tibble(data.frame(NAME = c("Winter", "Spring", "Summer", "Fall"),
                             SEASON = as.factor(1:4)))

# join DF with seasons reference dataframe
CumPrcp <- CumPrcp %>%
  inner_join(seasons, by = join_by("SEASON"))
  

# create separate dataframe that includes only the final month of each year
CumPrcpSeasons <- CumPrcp %>%
  group_by(YEAR, SEASON) %>%
  filter(sumPRCP == max(sumPRCP))
```

## Plot

```{r}
plot_CumPrcp <- ggplot() +
  # create lines for cumulative precipitation, grouped by year
  ## data was reordered so that the color attributes would align
  geom_path(data = CumPrcp[order(CumPrcp$YEAR, rev(CumPrcp$DATE)), ],
            aes(x = DATE, y = sumPRCP, group = YEAR, color = NAME),
            linewidth = 1) +
  # create connecting lines to show precip at the end of each season by year
  geom_path(data = CumPrcpSeasons,
            aes(x = DATE, y = sumPRCP, group = SEASON, color = NAME), 
            linewidth = 1, alpha = 0.5) +
  # modify x-axis breaks, labels, and margin
  scale_x_date(date_breaks = "2 year", date_labels = "%Y", 
               name = "Year", expand = c(0,400)) +
  # use custom palette, reorder season names in legend
  scale_color_manual(values = (pal2), 
                     breaks = c("Fall", "Summer", "Spring", "Winter")) +
  # relabel y-axis and titles
  labs(y = "Precipitation (mm)",
       title = "Cumulative Annual Precipitation",
       subtitle = "2000 - 2024 | Bellingham International Airport Weather Station") +
  # center titles
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) 

plot_CumPrcp
```

Fig. 2: Cumulative Annual Precipitation at Bellingham International Airport's weather station, 2000 to 2024. Colors indicate season. Horizontal lines show the cumulative precipitation at the end of each season across all years.

# Reflection

The biggest trouble I had in this assignment was getting the colors for geom_path to align correctly with the end of the season, instead of the second-to-last datapoint for each season. I found a reorder method that I used in the data call for the second plot, but despite knowing what that reorder did, I still don't quite fully grasp why it worked. 


